Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
tightening language in (2)

\def\ci{\perp\!\!\!\!\!\perp}

@@

\begin{abstract}
	Numerous causal discovery algorithms were developed to automatically learn
	directed acyclic graphs (DAGs) and other causal models from data. However,
	their adoption in applied domains remains limited, as researchers often
	prefer to construct DAGs manually based on domain knowledge. This
	preference arises due to several practical challenges with automated
	algorithms, such as their tendency to produce results that contradict
	obvious domain knowledge and their inability to distinguish Markov equivalent
	models. To assist researchers in constructing DAGs manually, we propose an iterative
	structure learning approach that combines domain knowledge with
	data-driven insights. Our method leverages conditional independence
	testing to iteratively identify variable pairs where an edge is
	either missing or superfluous. Based on this information, we can choose
	to add missing edges with appropriate orientation based on domain
	knowledge or remove unnecessary ones. We also give a method to rank
	these missing edges based on their impact on the overall model fit.
	In a simulation study, we find that this iterative approach to leverage domain 
	knowledge already starts outperforming purely data-driven structure learning if 
	the orientation of new edge is correctly determined in at least two out of three cases.
	We present a proof-of-concept implementation using a large language 
	model as a domain expert and a graphical user interface designed to 
	assist human experts with DAG construction.
\end{abstract}

@@

\begin{enumerate}
	\item \textbf{Lack of Trust:} While constraint-based algorithms are 
		often asymptotically
		consistent \citep{KalischB07}, they can and do make mistakes
		on  finite samples. These mistakes can be severe and contradict
		obvious domain knowledge (think of edges going into unmodifiable
		attributes such as Age). The choice of algorithm and hyperparameters
		significantly affects the output, making it
		difficult to assess reliability. Additionally, the absence of
		robust performance evaluation methods for any given dataset
		further reduce the confidence in their outputs. A recent paper
		therefore advised Epidemiologists to not attempt using structure
		learning algorithms without help of an expert \citep{Gururaghavendran_2024}.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, automated 
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. This structural uncertainty
		can make it difficult or impossible to apply the learned model for downstream
		tasks, such as identification or causal
		effect estimation \citep{Maathuis_2009,PerkovicTKM17}.
\end{enumerate}

@@

In this paper, we propose a structure learning method that leverages this CI
testing approach to assist researchers during manual DAG construction, rather
than just validating the model. Our approach follows an iterative
model-building strategy that integrates domain expertise with CI testing on
data. At each iteration, we use CI testing to identify pairs of variables that
lack (direct or indirect) connections or are connected by a superfluous edge.
 We then use this information
along with domain expertise to determine which changes to make.
The orientation of the edge when adding is determined based on the domain
expertise; critically, this does not require the domain expert to distinguish 
between direct and indirect effects. Our resulting approach can be 
thought of as a guidance during manual model construction that ensures the 
model remains consistent to the data.

@@

\section{Background}
\label{sec:background}
We denote random variables using uppercase letters, such as $ X $, and a set of
random variables by $ \bm{X} = \{X_1, \cdots, X_k\} $, where $ \rvert \bm{X}
\rvert = k $ indicates the number of variables in the set. We denote the
covariance between variables $ X $ and $ Y $ as $ \mathrm{cov}(X, Y) $, the
covariance matrix as $ \Sigma $, and the entry corresponding to covariance
between $ X $ and $ Y $ using $ \Sigma_{XY} $. In the context of this paper, we
consider causal DAGs that contain only observed variables. The structure of the
DAG is represented using a graph $ G = (V, E) $, where $ V $ are the nodes
corresponding to the random variables, and $ E $ are the edges denoting the
causal relationship between the variables. The set of parents of $ X $ in $ G $
are denoted as $ \textrm{Pa}_G(X) $, and its ancestors and descendants as $
\textrm{An}_G(X) $ and $ \textrm{De}_G(X) $ respectively. We denote a path, $
\pi(X, Y) = \{ X, V_0, V_1, \cdots, V_k, Y \} $ between $ X $ and $ Y $ in $ G
$ where each consecutive pair of variables in $ \pi $ are connected through an
edge. We say $ X $ and $ Y $ are d-connected in $ G $ given a conditioning set
$ \bm{Z} \subseteq V - \{X, Y\} $, if for a path $ \pi(X, Y) $ (i) for every
collider structure ($ V_1 \rightarrow V_2 \leftarrow V_3 $) on $ \pi(X, Y) $, $
\bm{Z} \cap \{ V_2, \textrm{De}_G(V_2) \} \ne \emptyset $, and (ii) for every
non-collider structure ($ V_1 \rightarrow V_2 \rightarrow V_3 $, $ V_1
\leftarrow V_2 \rightarrow V_3 $), $ V_2 \not \in \bm{Z} $. $ X $ and $ Y $ are
d-separated given $ \bm{Z} $ if no d-connecting path exist between them.

\subsection{Testing DAG Using CI Tests}
Our structure learning algorithm uses the idea of testing DAGs using CI tests.
In this section, we give an example of how this model testing works in
practice. We take the example of the MEC learned by the PC algorithm in
Figure~\ref{fig:intro}(b) using $ N=800 $. We first orient all the undirected
edges in the model based on our domain knowledge to obtain a DAG. In this case
we only have one undirected edge between \emph{Marital Status} and
\emph{Relationship} that we orient as \emph{Marital Status} $ \rightarrow $
\emph{Relationship} . We can then test the implied CIs of the DAG using our
dataset. Figure~\ref{fig:ci_table} shows the result of this testing on some of
the implied CIs. If we use a p-value threshold of $ 0.05 $, the first $ 2 $
implied CIs hold in the data whereas the remaining tests fail. One of the ways
to fix these failing tests is to add an edge between the variable pair $ X $
and $ Y $. While this method helps us in finding variable pairs where we need
to add an edge, it does not give us any information about the orientation of
this edge. We would have to rely on domain knowledge to determine this
orientation.

@@

\subsection{Effect Sizes for CI Tests}
\label{sec:ci_tests}

Our structure learning approach relies on using CI tests and a measure of
conditional association between variables -- also known as partial association.
We are interested in testing CI tests: $ X \ci Y \rvert \bm{Z} $ (with
potentially $ \bm{Z} = \emptyset $) in data and quantifying the strength of the
relationship between $ X $ and $ Y $ conditioned on $ \bm{Z} $. For the
quantification we use effect size measures based on the CI test. In this
section, we give an overview of some of the CI tests and effect size measures
for different types of (possibly mixed) data.

\paragraph{Both $ X $ and $ Y $ are continuous: }
When both $ X $ and $ Y $ are continuous, we can use a (partial) correlation
test for CI testing and Pearson's correlation coefficient can be used at the
effect size. When $ \bm{Z} = \emptyset $, the correlation coefficient is
defined as: $ r_{X, Y} = \mathrm{cov}(X, Y) / (\sigma_X \sigma_Y) $. When $
\bm{Z} \neq \emptyset $, partial correlation coefficient can be used. This is
estimated by fitting two regression models $ E_X: X \sim \bm{Z} $ and $ E_Y: Y
\sim \bm{Z} $, calculating the residuals $ R_X = X - E_X(\bm{Z}) $ and $ R_Y =
Y - E_Y(\bm{Y}) $, and then computing the correlation coefficient between the
residuals: $r_{X, Y \rvert \bm{Z}} = r_{R_X, R_Y}$

@@

Polyserial (when one is ordinal and the other is continuous) and Polychoric
(when both are ordinal) correlation have been used to estimate the covariance
matrix between them \citep{Poon1987}. Both methods make the assumption that the
observed ordinal variable is a result of thresholding a latent normally
distributed continuous variable. Under this assumption, the methods then try to
estimate the covariance matrix while maximizing the likelihood of the dataset.
Using the estimated covariance matrix, $ \Sigma $ we can perform a correlation
test for CI and compute the Pearson's correlation coefficient as the effect
size. When 
$\bm{Z} = \emptyset $, $ r_{X, Y} = \Sigma_{XY} / (\sqrt{\Sigma_{XX} \Sigma_{YY}}) $, 
	and when $ \bm{Z} \ne \emptyset $,
	$ r_{X, Y \rvert \bm{Z}} = - \Sigma^{-1}_{XY}/ (\sqrt{\Sigma^{-1}_{XX} \Sigma^{-1}_{YY}}) $.

\paragraph{All $ X $, $ Y $, and $ \bm{Z} $ are discrete (ordinal or categorical): }

In the case when we have a combination of ordinal and categorical variables, we
can use a residualization based CI test \citep{Ankan2023}. This test gives a
chi-square distributed test statistic, that can be used for computing an effect
size such as the Root Mean Squared Error of Approximation (RMSEA). Given a
chi-squared test statistic, $ \chi^2 $, with $ \textrm{df} $ degrees of
freedom, the RMSEA is defined as: $ \textrm{RMSEA}_{X, Y \rvert \bm{Z}} =
\sqrt{\chi^2 - \textrm{df}/ (\textrm{df} (N-1))} $, where $ N $ is the sample
size.

>>>>

\def\ci{\perp\!\!\!\!\perp}

@@

\begin{abstract}
	Numerous causal discovery algorithms were developed to automatically learn
	directed acyclic graphs (DAGs) and other causal models from data. However,
	their adoption in applied domains remains limited, as researchers often
	prefer to construct DAGs manually based on domain knowledge. This
	preference arises due to several practical challenges with automated
	algorithms, such as their tendency to produce results that contradict
	obvious domain knowledge and their inability to distinguish Markov equivalent
	models. To assist researchers in constructing DAGs manually, we propose an iterative
	structure learning approach that combines domain knowledge with
	data-driven insights. Our method leverages conditional independence
	testing to iteratively identify variable pairs where an edge is
	either missing or superfluous. Based on this information, we can choose
	to add missing edges with appropriate orientation based on domain
	knowledge or remove unnecessary ones. We also give a method to rank
	these missing edges based on their impact on the overall model fit.
	In a simulation study, we find that this iterative approach to leverage domain 
	knowledge already starts outperforming purely data-driven structure learning if 
	the orientation of new edge is correctly determined in at least one out of two cases.
	We present a proof-of-concept implementation using a large language 
	model as a domain expert and a graphical user interface designed to 
	assist human experts with DAG construction.
\end{abstract}

@@

\begin{enumerate}
	\item \textbf{Lack of Trust:} While constraint-based algorithms are 
		often asymptotically
		consistent \citep{KalischB07}, they can and do make mistakes
		on  finite samples. These mistakes can be severe and contradict
		obvious domain knowledge (think of edges going into unmodifiable
		attributes such as Age). The choice of algorithm and hyperparameters
		significantly aa=ffect the output, making it
		difficult to assess reliability. Additionally, the absence of
		robust performance evaluation methods for any given dataset
		further reduce the confidence in their outputs. A recent paper
		therefore advised Epidemiologists to not attempt using structure
		learning algorithms without help of an expert \citep{Gururaghavendran_2024}.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, automated 
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. This structural uncertainty
		can make difficult or impossible to apply the learned model for downstream
		tasks, such as identification or causal
		effect estimation \citep{Maathuis_2009,PerkovicTKM17}.
\end{enumerate}

@@

In this paper, we propose a structure learning method that leverages this CI
testing approach to assist researchers during manual DAG construction, rather
than just validating the model. Our approach follows an iterative
model-building strategy that integrates domain expertise with CI testing on
data. At each iteration, we use CI testing to identify pairs of variables that
lack (direct or indirect) connections or are connected by a superfluous edge.
 We then use this information
along with domain expertise to determine which changes.
The orientation of the edge when adding is determined based on the domain
expertise; critically, this does not require the domain expert to distinguish 
between direct and indirect effects. Our resulting approach can be 
thought of as a guidance during manual model construction that ensures the 
model remains consistent to the data.

@@

\section{Background}
\label{sec:background}
We denote random variables using uppercase letters like $X$, and a set of
random variables by $ \bm{X} = \{X_1, \cdots, X_k\} $, where $ \rvert \bm{X}
\rvert = k $ is the number of variables in the set. We denote the
covariance between variables $ X $ and $ Y $ as $ \mathrm{cov}(X, Y) $, a 
covariance matrix as $ \Sigma $, and the entry corresponding to the covariance
between $ X $ and $ Y $ as $ \Sigma_{XY} $. We consider causal DAGs that contain 
only observed variables. A DAG $ G = (V, E) $ is an acyclic directed graph
whose nodes $ V $ corresponding to random variables and whose edges $ E $ represent
direct causal relationships \cite{Pearl2009}. For an edge $e=X \to Y$, 
$\textrm{reverse}(e)=Y \to X$. The set of parents of $ X $ in $ G $
is denoted as $ \textrm{Pa}_G(X) $, and its ancestors and descendants as $
\textrm{An}_G(X) $ and $ \textrm{De}_G(X) $, respectively; we use the convention
that $X \notin \textrm{An}_G(X)$ and $X \notin \textrm{De}_G(X)$. We denote a path, $
\pi(X, Y) = \{ X, V_0, V_1, \cdots, V_k, Y \} $ between $ X $ and $ Y $ in $ G
$ where consecutive pairs of variables in $ \pi $ are connected by an
edge. We say $ X $ and $ Y $ are d-connected in $ G $ given a conditioning set
$ \bm{Z} \subseteq V - \{X, Y\} $, if for a path $ \pi(X, Y) $ (i) for every
collider structure ($ V_1 \rightarrow V_2 \leftarrow V_3 $) on $ \pi(X, Y) $, $
\bm{Z} \cap \{ V_2, \textrm{De}_G(V_2) \} \ne \emptyset $, and (ii) for every
non-collider structure ($ V_1 \rightarrow V_2 \rightarrow V_3 $, $ V_1
\leftarrow V_2 \rightarrow V_3 $), $ V_2 \not \in \bm{Z} $. $ X $ and $ Y $ are
d-separated given $ \bm{Z} $ if no d-connecting path exist between them.

\subsection{Conditional Independence Tests}
\label{sec:ci_tests}
Our structure learning algorithm uses the idea of testing DAGs using CI statements
like $ X \ci Y \rvert \bm{Z} $ (with
potentially $ \bm{Z} = \emptyset $).
For example, take the MEC learned by the PC algorithm in
Figure~\ref{fig:intro}(b) using $ N=800 $, and suppose that we orient 
the undirected edge between \emph{Marital Status} and
\emph{Relationship}  as \emph{Marital Status} $ \rightarrow $
\emph{Relationship}. We can then read implied CIs of the DAG using $d$-separation,
and test them in our dataset. Figure~\ref{fig:ci_table} shows the result of some tests. 
Using a significance threshold  $\alpha=0.05 $, the first $ 2 $
implied CIs hold in the data whereas the remaining tests fail. One of the ways
to fix these failing tests is to add an edge between the variable pair $ X $
and $ Y $. While this method helps us in finding variable pairs where we need
to add an edge, it does not give us any information about the orientation of
this edge. In this paper, we use domain knowledge to determine this
orientation.

@@

In addition to determining sigificance, we are interested in quantifying the 
\emph{strength} of any partial association between $X$ and $Y$ after conditioning 
on $\mathbf{Z}$. This metric depends on the type of CI test used. In the following,
we briefly overview some CI tests and effect size measures
for different types of (possibly mixed) data.

\paragraph{Both $ X $ and $ Y $ are continuous: }
When both $ X $ and $ Y $ are continuous, we can use a (partial) correlation
test for CI testing and Pearson's correlation coefficient can be used as the
effect size. When $ \bm{Z} = \emptyset $, the correlation coefficient is
defined as: $ r_{X, Y} = \mathrm{cov}(X, Y) / (\sigma_X \sigma_Y) $. When $
\bm{Z} \neq \emptyset $, the partial correlation coefficient can be used instead.
This is estimated by fitting two regression models $ E_X: X \sim \bm{Z} $ and $ E_Y: Y
\sim \bm{Z} $, calculating the residuals $ R_X = X - E_X(\bm{Z}) $ and $ R_Y =
Y - E_Y(\bm{Y}) $, and then computing the correlation coefficient between the
residuals: $r_{X, Y \rvert \bm{Z}} = r_{R_X, R_Y}$

@@

Polyserial (for continuous $Y$) and polychoric
(ordinal $Y$) correlations are used to estimate correlations 
involving ordinal variables \citep{Poon1987}. Both methods make the assumption that the
observed ordinal variable is a result of thresholding a latent normally
distributed continuous variable. Under this assumption, the methods then 
estimate the threshold values and covariance matrix using maximum likelihood.
Using the estimated covariance matrix, $ \Sigma $ we can perform a correlation
test for CI and compute the Pearson's correlation coefficient as the effect
size in the same way as for continuous $ X, Y $ and $ \bm{Z} $, i.e., 
$\bm{Z} = \emptyset $, $ r_{X, Y} = \Sigma_{XY} / (\sqrt{\Sigma_{XX} \Sigma_{YY}}) $, 
	and when $ \bm{Z} \ne \emptyset $,
	$ r_{X, Y \rvert \bm{Z}} = - \Sigma^{-1}_{XY}/ (\sqrt{\Sigma^{-1}_{XX} \Sigma^{-1}_{YY}}) $.

\paragraph{$ X $, $ Y $, and $ \bm{Z} $ are all discrete (ordinal or categorical): }

For combinations of ordinal and categorical variables, we
can use a residualization-based CI test \citep{Ankan2023} that returns a 
chi-square distributed test statistic. Given the statistic, $ \chi^2 $, with $ \textrm{df} $ degrees of
freedom, we use the Root Mean Squared Error of Approximation (RMSEA) defined as
 $ \textrm{RMSEA}_{X, Y \rvert \bm{Z}} =
\sqrt{\max(0,\chi^2 - \textrm{df})/ (\textrm{df} (N-1))} $, where $ N $ is the sample
size. This effect size can be used for any statistical test with a chi-square 
distributed test statistic.
